D3064R0: How to Avoid OOTA Without Really Trying

### 2024-02-23 ootavacuous.tex
Apply Alan Stern feedback on store-load commit order
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Computer systems are expected to continue increasing their use of
speculative execution.
Nevertheless, the main points of this paper will remain unaffected.
To see this, keep in mind that on real systems, observable effects
must eventually be committed, but that nothing can be committed
until all of the computations that determine those effects remain
speculative.
Furthermore, instructions take time to execute, even when executed
speculatively.
For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{y[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then line~11 can cause the value 42 to be loaded into \co{r1}, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

Therefore, nonspeculative loads are prohibited from loading values from
speculative stores.
And yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent all such
speculative OOTA cycles from ever committing.

As a special case of this prohibition, a value stored speculatively must
not be exposed to nonspeculative loads from other threads.
In the further special case of multiple hardware threads sharing a core,
one might imagine speculation involving multiple threads.
However, in this case, any squashing of speculation must squash the
full extent of that speculation, across all of the hardware threads
that were involved.

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to nonspeculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated nonspeculatively.
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete nonspeculatively.

One could imagine a conditional-store instruction, similar to existing
conditional-move instructions, but with condition controlling a store
to memory rather than a move to a register.
The condition would control whether or not the store was executed,
and prior conditional-move instructions could control the value
stored and the address stored to.
With or without prior conditional-move instructions, but assuming a
compiler doing single-thread analysis, a conditional-store instruction
permits generation of code corresponding to lines~10 and~11 of
Listing~\ref{lst:Causality Test Case 2}
on page~\pageref{lst:Causality Test Case 2}
without use of a conditional branch.
This means that the control dependency would extend from the loads on
lines~8 and~9 to the store on line~11 and nowhere else, potentially
enabling more hardware optimizations than would be permitted for code
relying on conditional branches.
Nevertheless, the conditional store could not be committed while any
the instructions it depends on were still being speculatively executed.

Finally, the prohibition of hardware OOTA applies within a single
multi-threaded core as well as between cores.
This means that for any set of speculative accesses to be committed,
all loads in that set must have been confirmed by corresponding
nonspeculative stores.

>>>>

Computer systems are expected to continue increasing their use of
speculative execution.
Nevertheless, the main points of this paper will remain unaffected.
To see this, keep in mind that on real systems, observable effects
must eventually be committed, but that nothing can be committed
until all of the computations that determine those effects have
committed.
Furthermore, instructions take time to execute, even when executed
speculatively.
For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{y[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~9 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then line~11 can cause the value 42 to be loaded into \co{r1}, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

Therefore, a given load is prohibited from committing until after the
store that it reads from has committed.
And yes, this does mean that hardware can produce OOTA cycles during
speculative execution, but the hardware is required to prevent all such
speculative OOTA cycles from ever committing.
As a special case of this prohibition, if a speculated load obtains
a value from a store that has not yet committed, then if that store
is squashed, the load must also be squashed, even if that load and
that store are executed by different CPUs.

Applying this to Listing~\ref{lst:OOTA Cycle}, lines~10 and~12 cannot be
committed until after the load in line~8 and the conditional on line~9
have both committed.
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 have committed.

One could imagine a conditional-store instruction, similar to existing
conditional-move instructions, but with condition controlling a store
to memory rather than a move to a register.
The condition would control whether or not the store was executed,
and prior conditional-move instructions could control the value
stored and the address stored to.
With or without prior conditional-move instructions, but assuming a
compiler doing single-thread analysis, a conditional-store instruction
permits generation of code corresponding to lines~10 and~11 of
Listing~\ref{lst:Causality Test Case 2}
on page~\pageref{lst:Causality Test Case 2}
without use of a conditional branch.
This means that the control dependency would extend from the loads on
lines~8 and~9 to the store on line~11 and nowhere else, potentially
enabling more hardware optimizations than would be permitted for code
relying on conditional branches.
Nevertheless, the conditional store could not be committed until all of
the instructions it depends on have committed.

Finally, the prohibition of hardware OOTA applies within a single
multi-threaded core as well as between cores.
This means that for any set of speculative accesses to be committed,
all loads in that set must have committed, which in turn requires that
all store providing the loads' values must also have committed.
